Conceptualizing open health data platforms for low- and middle income countries

### 2024-04-02 index.tex
Update render

\subsection{Open technologies: available digital public
goods}\label{open-technologies-available-digital-public-goods}

Many components of the OpenHIE specification are now available as a
digital public goods. Table~\ref{tbl-digital-public-goods} lists
components that are currently available for implementing the OpenHIE
framework using open source, digital public goods that are compliant
with the FHIR standard, illustrating the maturity of this ecosystem and
development community. With the launch of the Instant OpenHIE
configuration toolkit\citep{InstantOpenHIEv2}, it has become easier to
set up, explore and develop HIEs thereby reducing costs and skills
required for software developers to deploy an OpenHIE architecture for
quicker solution testing and as a starting point for faster production
implementation and customisation. Several frameworks are available that
offer a set of preconfigured components out of the box, such as for
example:

\begin{itemize}
\tightlist
\item
  the ``Open Smart Register Platform'' (OpenSRP), that focuses on
  providing a mobile-first platform, including a FHIR native app
  designed to support the WHO Smart Guidelines
\item
  the OpenHIM Platform, a reference implementation of the Instant
  OpenHIE framework, providing an easy way to set up, manage and operate
  various HIE configurations
\end{itemize}

@@

To evaluate the extended OpenHIE architecture described above, we first
consider the OpenHIM Platform. The Open Health Information Mediator
(OpenHIM, \href{OpenHIMhttp://openhim.org/}{http://openhim.org/}))
component is the reference implementation of the Interoperability Layer
(IOL) as defined in the OpenHIE specification. The most current version
(8.4.2 at the time of writing) provides all the core functions including
central point of access for the services of the HIE; routing functions;
central logging for auditing and debugging purposes; and
orchestration/mediation mechanisms to co-ordinate requests. This
reference implementation of the IOL has recently been extended as a
reference implementation of the so-called `Instant OpenHIE' concept,
that aims to address the primary needs of (i) allowing implementers to
engage with a preconfigured health information exchange solution and
running tools (based on the architecture) and test their applicability
and functionality with a real health context problem; and (ii) having a
packaged reference version of the OpenHIE architecture that is comprised
of a set of reference technologies and other appropriate tools that form
the building blocks of the health information exchange that can be
configured and extended to support particular use cases
\citep{digitalsquare2020instantopenhie}. The Instant HIE project allows
OpenHIE components to be packages up, deployed, operated and scaled via
a simple Command Line Interface (CLI, version 2 released in 2023
\url{https://github.com/openhie/instant-v2}). The OpenHIM Platform
(\url{https://jembi.gitbook.io/openhim-platform}) provides a set of
bundled packages (called recipes) that can be deployed using the Instant
OpenHIE deployment tool. In the following we will evaluate the recipe
for ``a central data repository with a data warehouse'' that provides
``A FHIR-based Shared Health record linked to a Master Patient Index
(MPI) for linking and mathing patient demographics and a default
reporting pipeline to transform and visualise FHIR data''
(\url{https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing}).

@@

\caption{\label{fig-openhim-platform}Overview of the default data stack
of the OpenHIM Platform. The default stack (top, darkblue) consists of
Kafka, Clickhouse and Superset. An alternative solution based on the ELK
stack is also supported (bottom, lightblue), consisting of
Elasticsearch, Logstash and Kibana.}

@@

Figure~\ref{fig-openhim-platform} shows a schematic overview of two data
stacks that are supported in the OpenHIM platform. The Shared Health
Record (SHR, implemented with HAPI FHIR server) and the Client Registry
(CR, implemented with JeMPI server) are the sources that store clinical
FHIR data and patient demographic data, respectively. The default data
stack is based on streaming ingestion using Kafka into a Clickhouse
database. As part of the ingestion, incoming FHIR bundles that contain
multiple FHIR resources are unbundled in separate topics. Subsequently,
each FHIR resource topic is flatted using mappers Unbundling and
flattening of FHIR resources is done in Kafka using FHIRPath. Superset
is used as the tool for consuming the data to create dashboard
visualizations.

The OpenHIM platform also support data analytics based on the ELK stack,
where data is ingested in bulk using Logstash, stored in Elasticsearch
and made available for consumption in Kibana. Also here, the incoming
FHIR bundles are unbundled in Logstash into separate FHIR resources.
However, given that Elasticsearch is a document-based search engine, the
FHIR resources are stored as-is with no flattening. Exploring and
analysing the data requires writing queries in Elasticsearch Query
Language (ES\textbar QL), either through the query interface of
Elasticsearch or using Kibana.

@@

MomCare was launched in Kenya
\citep{huisman2022digital, sanctis2022maintaining} and Tanzania
\citep{shija2021access, mrema2021application} in 2017 and 2019
respectively, with the objective to create transparency on the journeys
of pregnant mothers. The programme is built on three pillars: journey
tracking, quality support and a mobile wallet. MomCare distinguishes two
user groups: mothers are supported during their pregnancy through
reminders and surveys, using SMS as the digital mode of engagement.
Health workers are equipped with an Android-based application, in which
visits, care activities and clinical observations are recorded.
Reimbursements of the maternal clinic are based on the data captured
with SMS and the app, thereby creating a conditional payment scheme,
where providers are partially reimbursed up-front for a fixed bundle of
activities, supplemented by bonus payments based on a predefined set of
care activities.

In its original form, the MomCare programme uses predominantly closed
digital platforms. In Kenya, M-TIBA is the primary digital platform, on
top of which a relatively lightweight custom app has been built as the
engagement layer for the health workers \citep{huisman2022digital}.
M-TIBA provides data access through its data warehouse platform for the
MomCare programme, however, this is not a standardized, general purpose
API. In the case of Tanzania, a stand-alone custom app is used which
does not provide an interface of any kind for interacting with the
platform \citep{mrema2021application}. Given these constraints, to date
the MomCare programme uses a custom-built data warehouse environment as
its main data platform, on which data extractions, transformations and
analysis are performed to generate the operational reports. Feedback
reports for the health workers, in the form of operational dashboards,
are made accessible through the app. Similar reports are provided to the
back-office for the periodic reimbursement to the clinics.

In the periode 2022-2023, a new iteration of MomCare \ldots One of the
practical objectives of the DataCare project was to assess the
complexity and effort required to implement the data transformation
scripts, and reproduce the existing analytic reports. The data was
transformed into 10 FHIR v4 resources as listed in table 1 with the
number of records per resource type. The conceptual data model of the
existing MomCare app could readily be transformed into the FHIR standard
using SQL and validated with a Python library \citep{islam2023fhir}.

The largest challenge during the transformation process pertained to the
absence of unique business identifiers for patients and healthcare
organizations. For patients, either the mobile phone number or the
healthcare insurance number was taken, depending on availability. A
combination of name, address and latitude/longitude coordinates were
used to uniquely identify organizations and locations, as Tanzania does
not have a system in place for this purpose.

The transformed and validated data is uploaded into the FHIR server on a
daily basis using an automated cloud function. Analysis of bulk data was
done by directly reading the standard newline delimited JSON into the
Python pandas data analysis library. Cross checking the output with
queries on the original data confirmed that the whole data pipeline
produced consistent results. For example, the report of the antenatal
coverage metric (number of pregnancies with four or more visits) could
be reproduced per patient journey and aggregated (per year, per
organization etc.) as required for the MomCare reports.

@@

\begin{itemize}
\tightlist
\item
  Object of openness: what data-related resources should data platforms
  make available when opening up (e.g.~data, data products, datadriven
  insights, analytics modules)? Which user groups derive value from
  accessing data-related resources from data platforms (e.g.~data
  providers, data users, intermediaries, developers)?
\item
  Unit of analysis: what is platform-to-platform openness in the context
  of data platforms, given the expectation that different HDPs will
  emerge at various aggregation levels? How do we distinguish
  meta-platforms, forking, and platform interoperability?
\item
  Risk of openness: What are the novel (negative) implications of
  opening up data platforms? How can reflexivity in design help
  providers to resolve the negative implications of openness?
\end{itemize}

>>>>

\subsection{Open technologies: deploying Instant OpenHIE with digital
public
goods}\label{open-technologies-deploying-instant-openhie-with-digital-public-goods}

Today, many components of the OpenHIE specification are now available as
a digital public goods, as listed in
Table~\ref{tbl-digital-public-goods}. To further ease the development,
configuration and deployment of health information exchanges, the
concept of `Instant OpenHIE' has been championed to (i) allow
implementers to engage with a preconfigured health information exchange
solution and running tools (based on the architecture) and test their
applicability and functionality with a real health context problem; and
(ii) have a packaged reference version of the OpenHIE architecture that
is comprised of a set of reference technologies and other appropriate
tools that form the building blocks of the health information exchange
that can be configured and extended to support particular use cases
\citep{InstantOpenHIEv2}. Besides the core functional components of the
OpenHIE architecture, the Instant OpenHIE toolkit allows packaging and
integration of generic components such as Identity and Access Managment
(IAM) and a reverse proxy gateway. In the following, we will evaluatie
three of such configurations, with the aim to conceptualize and evaluate
the proposed Data and Analytics Services domain of of the OpenHIE
architecture.

@@

To evaluate the extended OpenHIE architecture described above, we first
consider the OpenHIM Platform. The Open Health Information Mediator
(OpenHIM, \href{OpenHIMhttp://openhim.org/}{http://openhim.org/}))
component is the reference implementation of the Interoperability Layer
(IOL) as defined in the OpenHIE specification. The most current version
(8.4.2 at the time of writing) provides all the core functions including
central point of access for the services of the HIE; routing functions;
central logging for auditing and debugging purposes; and
orchestration/mediation mechanisms to co-ordinate requests. By
extension, the OpenHIM Platform
(\url{https://jembi.gitbook.io/openhim-platform}) is a reference
implementation of a set of Instant OpenHIE configurations, refered to as
`recipes' in the documentation. In the following we will evaluate the
recipe for ``a central data repository with a data warehouse'' that
provides ``A FHIR-based Shared Health record linked to a Master Patient
Index (MPI) for linking and mathing patient demographics and a default
reporting pipeline to transform and visualise FHIR data''
(\url{https://jembi.gitbook.io/openhim-platform/recipes/central-data-repository-with-data-warehousing}).

@@

\caption{\label{fig-openhim-platform}Overview of the default data stack
of the OpenHIM Platform. The default stack (top, red) consists of Kafka,
Clickhouse and Superset. An alternative solution based on the ELK stack
is also supported (bottom, orange), consisting of Elasticsearch,
Logstash and Kibana.}

@@

Figure~\ref{fig-openhim-platform} shows a schematic overview of two data
stacks that are supported in the OpenHIM platform. The Shared Health
Record (SHR, implemented with HAPI FHIR server) and the Client Registry
(CR, implemented with JeMPI server) are the sources that store clinical
FHIR data and patient demographic data, respectively. The default data
stack is based on streaming ingestion using Kafka into a Clickhouse
database. As part of the ingestion, incoming FHIR bundles that contain
multiple FHIR resources are unbundled in separate topics using a generic
Kafka utility component. Subsequently, each FHIR resource topic is
flatted with Kafka mappers that use FHIRPath. Superset is used as the
tool for consuming the data to create dashboard visualizations.

The OpenHIM platform also support data and analytics based on the ELK
stack, where data is ingested in bulk using Logstash, stored in
Elasticsearch and made available for consumption in Kibana. Also here,
the incoming FHIR bundles are unbundled in Logstash into separate FHIR
resources. However, given that Elasticsearch is a document-based search
engine, the FHIR resources are stored as-is with no flattening.
Exploring and analysing the data requires writing queries in
Elasticsearch Query Language (ES\textbar QL), either through the query
interface of Elasticsearch or using Kibana.

@@

MomCare was launched in Kenya
\citep{huisman2022digital, sanctis2022maintaining} and Tanzania
\citep{shija2021access, mrema2021application} in 2017 and 2019
respectively, with the objective to improve health outcomes for maternal
and antenatal care. MomCare distinguishes two user groups: mothers are
supported during their pregnancy through reminders and surveys, using
SMS as the digital mode of engagement. Health workers are equipped with
an Android-based application, in which visits, care activities and
clinical observations are recorded. Reimbursements of the maternal
clinic are based on the data captured with SMS and the app, thereby
creating a conditional payment scheme, where providers are partially
reimbursed up-front for a fixed bundle of activities, supplemented by
bonus payments based on a predefined set of care activities.

In its original form, the MomCare programme used closed digital
platforms. In Kenya, M-TIBA is the primary digital platform, on top of
which a relatively lightweight custom app has been built as the
engagement layer for the health workers \citep{huisman2022digital}.
M-TIBA provides data access through its data warehouse platform for the
MomCare programme, however, this is not a standardized, general purpose
API. In the case of Tanzania, a stand-alone custom app is used which
does not provide an interface of any kind for interacting with the
platform \citep{mrema2021application}. Given these constraints, the
first iteration of the MomCare programme used a custom-built data
warehouse environment as its main data platform, on which data
extractions, transformations and analysis are performed to generate the
operational reports. Feedback reports for the health workers, in the
form of operational dashboards, are made accessible through the app.
Similar reports are provided to the back-office for the periodic
reimbursement to the clinics.

Clearly, a more open and scaleable platform was required if MomCare was
to be implemented in more regions. This need led to a redesign of the
underlying technical infrastructure of the MomCare project. The
objectives of this work were in fact to demonstrate a solution design
that could support the first three types of data sharing listed in
Table~\ref{tbl-types-data-sharing}. First, to investigate the viability
of using FHIR for bulk data sharing, MomCare Tanzania was used a testbed
to assess the complexity and effort required to implement the facade
pattern to integrate the legacy system into the FHIR data standard.
Using the longitudinal dataset from approximately 28 thousand patient
records, FHIR transformations script were developed and deployed using
the mediator function of the IOL. The data was transformed into 10 FHIR
v4 resources and the conceptual data model of the existing MomCare app
could readily be transformed into the FHIR standard using SQL and
validated with a Python library \citep{islam2023fhir}. The largest
challenge during the transformation process pertained to the absence of
unique business identifiers for patients and healthcare organizations.
For patients, either the mobile phone number or the healthcare insurance
number was taken, depending on availability. A combination of name,
address and latitude/longitude coordinates were used to uniquely
identify organizations and locations, as Tanzania does not have a system
in place for this purpose.

The second objective was to reproduce existing analytic reports, using
the bulk FHIR data format as input. Here, the focus was to standardize
the logic required for producing metrics and reports. The transformed
and validated data is uploaded into the FHIR server on a daily basis
using an automated cloud function. Analysis of bulk data was done by
directly reading the standard newline delimited JSON into the Python
pandas data analysis library. Cross checking the output with queries on
the original data confirmed that the whole data pipeline produced
consistent results. For example, the report of the antenatal coverage
metric (number of pregnancies with four or more visits) could be
reproduced per patient journey and aggregated (per year, per
organization etc.) as required for the MomCare reports.

TO DO: explain logic of patient-timeline table. Write standard
transformation to go from FHIR resources to this standard table. On top
of that the actual metrics and reporting. Explain serverless: we wanted
to get rid of resource-heavy data visualization tools. This led to the
idea of serverless: using duckdb-wasm and pipelines of cloud functions.

The third objective was to run a technical feasibility test for
federated analytics. Using the MPC platform of Roseman Labs, we managed
to do aggregations in the blind \ldots{} TO DO: explain that we managed
to reproduce the reports we generated in the clear, but then in the
blind. Note, however, that in the remainder we will focus on first two
types of data sharing.

Based on these experiments, we arrived at the following design for the
data \& analytics services

\begin{itemize}
\tightlist
\item
  Use `serverless' file-based storage: bulk copy of data as-is in
  parquet

  \begin{itemize}
  \tightlist
  \item
    Tension: how to manage change data capture
  \item
    Tension: how to manage access rights
  \end{itemize}
\item
  Use SQL-on-FHIR-v2 to create tabular views.

  \begin{itemize}
  \tightlist
  \item
    Example: patient timeline
  \item
    TO DO: rewrite patient timeline queries with SQL-on-FHIR-v2 and run
    it with Pathling
  \end{itemize}
\item
  Use semantic modeling layer to define metrics

  \begin{itemize}
  \tightlist
  \item
    There are many options: dbt, cube.dev
  \item
    Fulfills same function as ADX/mADX IHE profile in OpenHIE
    specification
  \item
    Tension: going from patient-timeline to reported metrics still isn't
    standardized. This is where Ibis/Substrait comes in. Substrait as IR
    for cross-language serialization for relational algebra. Can be
    executed on different backends. Write once, run on different
    engines.
  \end{itemize}
\item
  Distribute and publish reports on resource-constrained devices

  \begin{itemize}
  \tightlist
  \item
    duckdb
  \item
    sveltekit
  \end{itemize}
\end{itemize}

TO DO: Add diagram

@@

\begin{itemize}
\item
  Object of openness: what data-related resources should data platforms
  make available when opening up (e.g.~data, data products, datadriven
  insights, analytics modules)? Which user groups derive value from
  accessing data-related resources from data platforms (e.g.~data
  providers, data users, intermediaries, developers)?
\item
  Unit of analysis: what is platform-to-platform openness in the context
  of data platforms, given the expectation that different HDPs will
  emerge at various aggregation levels? How do we distinguish
  meta-platforms, forking, and platform interoperability?
\item
  Risk of openness: What are the novel (negative) implications of
  opening up data platforms? How can reflexivity in design help
  providers to resolve the negative implications of openness?
\item
  Answers/insights to above:

  \begin{itemize}
  \tightlist
  \item
    Openness of standardized view on FHIR data and cross-language
    serialization of relational algebra makes it possible to fully
    standardize the workflow from start to finish
  \item
    Platform-to-platform: MPC
  \item
    Risk of openness: difficult to answer \ldots{}
  \end{itemize}
\end{itemize}
